---
title: "01 — Exploratory Data Analysis"
author: "PSTAT 134 — AI Writing Detector"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    theme: simplex
    code-fold: show
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
here::i_am("notebooks/01-eda.qmd")

source(here::here("R", "data.R"))
source(here::here("R", "features.R"))

library(tidyverse)
library(scales)

theme_set(theme_minimal(base_size = 12))
```

## Load the data

```{r}
#| label: load
df <- load_essays(here::here("data", "raw", "AI_Human.csv"))
dim(df)
```

```{r}
head(df)
```

## Class balance

The dataset is intentionally balanced — confirm the 50/50 split is preserved.

```{r}
#| label: class-balance
df |>
  count(generated) |>
  mutate(pct = n / sum(n))
```

## Length distributions

```{r}
#| label: lengths
df_len <- df |>
  mutate(
    n_chars = nchar(text),
    n_words = lengths(stringr::str_extract_all(text, "\\b\\w+\\b"))
  )

ggplot(df_len, aes(x = n_words, fill = generated)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.6) +
  scale_x_continuous(labels = comma) +
  labs(title = "Word count by class", x = "words", y = "essays")
```

```{r}
ggplot(df_len, aes(x = generated, y = n_words, fill = generated)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(title = "Word count, boxplot", x = NULL, y = "words")
```

## Stylometric & readability features

The teammate plan calls for readability indices (Flesch-Kincaid, etc.) in
addition to the simple hand-crafted features, so `stylometric_features()` is
called with `include_readability = TRUE`.

```{r}
#| label: stylometric
style <- stylometric_features(df$text, include_readability = TRUE) |>
  mutate(generated = df$generated)

style |>
  group_by(generated) |>
  summarise(across(where(is.numeric), mean), .groups = "drop") |>
  pivot_longer(-generated, names_to = "feature", values_to = "mean") |>
  pivot_wider(names_from = generated, values_from = mean)
```

```{r}
#| label: stylometric-plot
#| fig-width: 12
#| fig-height: 7
style |>
  select(generated, avg_sentence_len_words, type_token_ratio, punct_rate,
         Flesch, Flesch.Kincaid, ARI) |>
  pivot_longer(-generated, names_to = "feature", values_to = "value") |>
  ggplot(aes(x = generated, y = value, fill = generated)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~ feature, scales = "free_y") +
  labs(x = NULL, y = NULL)
```

## Notes for the team

- Are the two classes separable on simple length / punctuation features alone?
- Which features look most promising as descriptive variables for the report?
- Anything weird (extremely short essays, duplicated text, encoding issues) we should clean before modeling?
