---
title: "03 — Baseline Models"
author: "PSTAT 134 — AI Writing Detector"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    theme: simplex
    code-fold: show
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
here::i_am("notebooks/03-baseline-models.qmd")

source(here::here("R", "features.R"))
source(here::here("R", "evaluation.R"))

library(tidyverse)
library(tidymodels)
library(textrecipes)
library(discrim)

tidymodels_prefer()
```

## Load processed training data

```{r}
train_df <- readr::read_csv(
  here::here("data", "processed", "train.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))

dim(train_df)
```

## Shared CV folds

Every baseline uses the same stratified 5-fold split so the comparison is fair.

```{r}
folds <- cv_folds(train_df)
folds
```

## Baseline 1: TF-IDF + Logistic Regression

The TF-IDF step lives inside the recipe, so the vocabulary is learned only
from the training fold during cross-validation. This avoids leaking test-fold
vocabulary into training.

```{r}
rec_tfidf <- tfidf_recipe(generated ~ text, data = train_df)

logreg_spec <- logistic_reg(penalty = 0.01, mixture = 0) |>
  set_engine("glmnet") |>
  set_mode("classification")

logreg_wf <- workflow() |>
  add_recipe(rec_tfidf) |>
  add_model(logreg_spec)

logreg_cv <- cv_summary(logreg_wf, folds)
logreg_cv
```

## Baseline 2: TF-IDF + Naive Bayes

```{r}
nb_spec <- naive_Bayes() |>
  set_engine("naivebayes") |>
  set_mode("classification")

nb_wf <- workflow() |>
  add_recipe(rec_tfidf) |>
  add_model(nb_spec)

nb_cv <- cv_summary(nb_wf, folds)
nb_cv
```

## Side-by-side comparison

```{r}
bind_rows(
  logreg_cv |> mutate(model = "logreg"),
  nb_cv     |> mutate(model = "naive_bayes")
) |>
  pivot_wider(id_cols = metric, names_from = model, values_from = mean)
```

## Notes

- These numbers are 5-fold CV means on the **training** set only.
- The test set is still untouched — final numbers are reported in notebook 04.
- If one baseline is dramatically ahead, note it here and bring it forward as
  the model to tune.
