---
title: "04 — Model Comparison, Tuning & Final Evaluation"
author: "PSTAT 134 — AI Writing Detector"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    theme: simplex
    code-fold: show
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
here::i_am("notebooks/04-model-comparison.qmd")

source(here::here("R", "features.R"))
source(here::here("R", "evaluation.R"))

library(tidyverse)
library(tidymodels)
library(textrecipes)
library(discrim)
library(vip)
library(kernlab)

tidymodels_prefer()
```

## Load processed train / test data

```{r}
train_df <- readr::read_csv(
  here::here("data", "processed", "train.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))

test_df <- readr::read_csv(
  here::here("data", "processed", "test.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))
```

## Compare candidate models with shared CV folds

Every workflow uses the **same** stratified folds and the **same** TF-IDF
recipe, so the comparison is fair.

```{r}
folds <- cv_folds(train_df)
rec_tfidf <- tfidf_recipe(generated ~ text, data = train_df, max_tokens = 1000)

specs <- list(
  logreg        = logistic_reg(penalty = 0.01, mixture = 0)         |> set_engine("glmnet")    |> set_mode("classification"),
  naive_bayes   = naive_Bayes()                                     |> set_engine("naivebayes")|> set_mode("classification"),
  linear_svm    = svm_linear(cost = 1)                                      |> set_engine("LiblineaR") |> set_mode("classification"),
  random_forest = rand_forest(trees = 500)                          |> set_engine("ranger")    |> set_mode("classification"),
  xgboost       = boost_tree(trees = 500, learn_rate = 0.05)        |> set_engine("xgboost")   |> set_mode("classification")
)

models_dir <- here::here("models")
dir.create(models_dir, showWarnings = FALSE, recursive = TRUE)
results_path <- here::here("models", "cv_results.rds")

# Load existing results if they exist, otherwise start fresh
if (file.exists(results_path)) {
  results <- readRDS(results_path)
  cat("Loaded existing results for:", paste(unique(results$model), collapse = ", "), "\n")
} else {
  results <- tibble()
}

# Only run models that haven't been completed yet
remaining <- setdiff(names(specs), unique(results$model))
cat("Models remaining:", paste(remaining, collapse = ", "), "\n")

for (name in remaining) {
  cat("\n--- Starting:", name, "at", format(Sys.time(), "%H:%M:%S"), "---\n")
  
  spec <- specs[[name]]
  wf <- workflow() |> add_recipe(rec_tfidf) |> add_model(spec)
  
  tryCatch({
    # SVM can't do ROC-AUC, use accuracy and F1 only
    if (name == "linear_svm") {
      res <- cv_summary(wf, folds, metrics = metric_set(accuracy, f_meas)) |> 
        mutate(model = name)
    } else {
      res <- cv_summary(wf, folds) |> mutate(model = name)
    }
    results <- bind_rows(results, res)
    saveRDS(results, results_path)
    cat("✓ Finished:", name, "at", format(Sys.time(), "%H:%M:%S"), "\n")
    print(res |> select(metric, mean, std_err))
  }, error = function(e) {
    cat("✗ Failed:", name, "\n")
    cat("  Error:", conditionMessage(e), "\n")
  })
}

# Final comparison table
results |>
  select(model, metric, mean, std_err) |>
  pivot_wider(id_cols = metric, names_from = model, values_from = mean)
```
Linear SMV model was ran without calculating ROC-AUC due to LiblineaR engine limitation. Re-run with AUC-ROC is not necessary as it falls below both Random Forest and XGBoost in both accuracy and F1. 

XGBoost has performed the best of the 5 models in all metrics. This will be tuned to find the best overall model.

## Pick the best model and tune it

Tune XGBoost model

```{r}
BEST_NAME <- "xgboost"

tune_spec <- boost_tree(trees = 500, learn_rate = tune(), tree_depth = tune()) |>
  set_engine("xgboost") |>
  set_mode("classification")

tune_wf <- workflow() |>
  add_recipe(rec_tfidf) |>
  add_model(tune_spec)

xgb_grid <- grid_regular(learn_rate(range = c(-3, -1)),
                         tree_depth(range = c(3,12)),
                         levels = 5)

tune_res_path <- here::here("models", "tune_res.rds")

if (file.exists(tune_res_path)) {
  tune_res <- readRDS(tune_res_path)
  cat("Loaded saved tuning results\n")
} else {
  tune_res <- tune::tune_grid(
    tune_wf,
    resamples = folds,
    grid      = xgb_grid,
    metrics   = cv_metrics(),
    control   = control_grid(event_level = "second")
  )
  saveRDS(tune_res, tune_res_path)
  cat("Tuning complete and saved\n")
}

best_params <- select_best(tune_res, metric = "roc_auc")
best_params
```
The best performing model was at a tree depth of 12 (max of range tested). We can look at a plot that shows how ROC AUC changes as tree depth increase to check if we need to rerun the model with a wider range.

```{r}
tune_res |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  ggplot(aes(x = tree_depth, y = mean, color = factor(learn_rate), group = factor(learn_rate))) +
  geom_line() +
  geom_point() +
  labs(title = "XGBoost tuning: ROC_AUC by tree depth and learn rate",
       x = "Tree Depth", y = "CV ROC AUC", color = "Learn Rate")
```
Plot shows that ROC AUC starts to plateau at a tree depth of 12. Rerun isn't necessary for risk of over-fitting.

```{r}
final_wf <- finalize_workflow(tune_wf, best_params)
final_fit <- fit(final_wf, data = train_df)
```

```{r}
final_fit_path <- here::here("models", "final_fit.rds")
saveRDS(final_fit, final_fit_path)
```


## Final evaluation on the held-out test set

This is the **only** time we touch the test set.

```{r}
report <- test_set_report(final_fit, test_df)
report$metrics
```

```{r}
report$confusion_matrix
```

```{r}
#| label: confusion-matrix
fig_dir <- here::here("reports", "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

cm_plot <- autoplot(report$confusion_matrix, type = "heatmap") +
  ggplot2::labs(title = "Confusion matrix — held-out test set")

ggplot2::ggsave(
  filename = file.path(fig_dir, "confusion_matrix.png"),
  plot = cm_plot, width = 5, height = 4, dpi = 150
)

cm_plot
```

## Inspect the most informative features

```{r}
#| label: top-features
final_fit |>
  extract_fit_parsnip() |>
  vip::vi() |>
  slice_max(abs(Importance), n = 30) |>
  arrange(desc(Importance))
```

## Persist the final model

```{r}
models_dir <- here::here("models")
dir.create(models_dir, showWarnings = FALSE, recursive = TRUE)

saveRDS(final_fit, file = file.path(models_dir, paste0(BEST_NAME, "_final.rds")))
```

## Notes for the report

- Compare the CV ranking to the test-set ranking — large gaps suggest
  overfitting to the CV folds.
- Use the top features and confusion matrix to argue *why* the model is making
  the predictions it makes.
- Discuss how well any of this would generalize to AI text from a different
  generator (e.g., a newer model than the one in the dataset).
