---
title: "04 — Model Comparison, Tuning & Final Evaluation"
author: "PSTAT 134 — AI Writing Detector"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    theme: simplex
    code-fold: show
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
here::i_am("notebooks/04-model-comparison.qmd")

source(here::here("R", "features.R"))
source(here::here("R", "evaluation.R"))

library(tidyverse)
library(tidymodels)
library(textrecipes)
library(discrim)
library(vip)

tidymodels_prefer()
```

## Load processed train / test data

```{r}
train_df <- readr::read_csv(
  here::here("data", "processed", "train.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))

test_df <- readr::read_csv(
  here::here("data", "processed", "test.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))
```

## Compare candidate models with shared CV folds

Every workflow uses the **same** stratified folds and the **same** TF-IDF
recipe, so the comparison is fair.

```{r}
folds <- cv_folds(train_df)
rec_tfidf <- tfidf_recipe(generated ~ text, data = train_df)

specs <- list(
  logreg      = logistic_reg(penalty = 0.01, mixture = 0)         |> set_engine("glmnet")    |> set_mode("classification"),
  naive_bayes = naive_Bayes()                                     |> set_engine("naivebayes")|> set_mode("classification"),
  linear_svm  = svm_linear()                                      |> set_engine("LiblineaR") |> set_mode("classification"),
  xgboost     = boost_tree(trees = 500, learn_rate = 0.05)        |> set_engine("xgboost")   |> set_mode("classification")
)

results <- imap_dfr(specs, function(spec, name) {
  wf <- workflow() |> add_recipe(rec_tfidf) |> add_model(spec)
  cv_summary(wf, folds) |> mutate(model = name)
})

results |>
  select(model, metric, mean, std_err) |>
  pivot_wider(id_cols = metric, names_from = model, values_from = mean)
```

## Pick the best model and tune it

Edit `BEST_NAME` once you know which candidate to take forward. The example
below tunes `penalty` for the regularized logistic regression.

```{r}
BEST_NAME <- "logreg"

tune_spec <- logistic_reg(penalty = tune(), mixture = 0) |>
  set_engine("glmnet") |>
  set_mode("classification")

tune_wf <- workflow() |>
  add_recipe(rec_tfidf) |>
  add_model(tune_spec)

penalty_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 8)

tune_res <- tune::tune_grid(
  tune_wf,
  resamples = folds,
  grid      = penalty_grid,
  metrics   = cv_metrics(),
  control   = control_grid(event_level = "second")
)

best_params <- select_best(tune_res, metric = "f_meas")
best_params
```

```{r}
final_wf <- finalize_workflow(tune_wf, best_params)
final_fit <- fit(final_wf, data = train_df)
```

## Final evaluation on the held-out test set

This is the **only** time we touch the test set.

```{r}
report <- test_set_report(final_fit, test_df)
report$metrics
```

```{r}
report$confusion_matrix
```

```{r}
#| label: confusion-matrix
fig_dir <- here::here("reports", "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

cm_plot <- autoplot(report$confusion_matrix, type = "heatmap") +
  ggplot2::labs(title = "Confusion matrix — held-out test set")

ggplot2::ggsave(
  filename = file.path(fig_dir, "confusion_matrix.png"),
  plot = cm_plot, width = 5, height = 4, dpi = 150
)

cm_plot
```

## Inspect the most informative features

```{r}
#| label: top-features
final_fit |>
  extract_fit_parsnip() |>
  vip::vi() |>
  slice_max(abs(Importance), n = 30) |>
  arrange(desc(Importance))
```

## Persist the final model

```{r}
models_dir <- here::here("models")
dir.create(models_dir, showWarnings = FALSE, recursive = TRUE)

saveRDS(final_fit, file = file.path(models_dir, paste0(BEST_NAME, "_final.rds")))
```

## Notes for the report

- Compare the CV ranking to the test-set ranking — large gaps suggest
  overfitting to the CV folds.
- Use the top features and confusion matrix to argue *why* the model is making
  the predictions it makes.
- Discuss how well any of this would generalize to AI text from a different
  generator (e.g., a newer model than the one in the dataset).
