---
title: "05 — Hyperparameter Tuning & Final Model"
author: "Nathan Kim — PSTAT 134 — AI Writing Detector"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    theme: simplex
    code-fold: show
execute:
  warning: false
  message: false
---

This notebook covers **Section 6** of the project outline: tuning the
best-performing model from the comparison in `04-model-comparison.qmd`
(XGBoost), tuning the **TF-IDF parameters alongside the model parameters**,
evaluating the final model on the held-out test set, and analyzing the
confusion matrix.

```{r}
#| label: setup
here::i_am("notebooks/05-tuning-final-model.qmd")

source(here::here("R", "features.R"))
source(here::here("R", "evaluation.R"))

library(tidyverse)
library(tidymodels)
library(textrecipes)

tidymodels_prefer()
set.seed(42)
```

## Load the processed train / test splits

These were produced in `02-preprocessing.qmd` and are committed under
`data/processed/`. The test set is **only** touched once, at the very end.

```{r}
#| label: load
train_df <- readr::read_csv(
  here::here("data", "processed", "train.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))

test_df <- readr::read_csv(
  here::here("data", "processed", "test.csv"),
  show_col_types = FALSE
) |>
  mutate(generated = factor(generated, levels = c("human", "ai")))

c(train = nrow(train_df), test = nrow(test_df))
```

## A tunable workflow (TF-IDF parameters + model parameters)

The model comparison fixed the recipe (`max_tokens = 1000`, default n-grams)
and only varied the XGBoost parameters. Section 6 calls for tuning the TF-IDF
representation jointly with the model, so here both sets of knobs carry
`tune()`:

- **TF-IDF:** `max_tokens` (vocabulary size) and `num_tokens` (max n-gram length).
- **XGBoost:** `learn_rate` and `tree_depth` (`trees` fixed at 500, matching the comparison).

Keeping the TF-IDF steps inside the recipe means the vocabulary is still
learned only from each analysis fold — no leakage during tuning.

```{r}
#| label: tunable-workflow
tune_rec <- recipe(generated ~ text, data = train_df) |>
  step_tokenize(text) |>
  step_stopwords(text) |>
  step_ngram(text, num_tokens = tune(), min_num_tokens = 1L) |>
  step_tokenfilter(text, max_tokens = tune(), min_times = 2L) |>
  step_tfidf(text)

xgb_spec <- boost_tree(
  trees      = 300,
  learn_rate = tune(),
  tree_depth = tune()
) |>
  set_engine("xgboost", nthread = parallel::detectCores()) |>
  set_mode("classification")

tune_wf <- workflow() |>
  add_recipe(tune_rec) |>
  add_model(xgb_spec)
```

## Search space

We use a randomized search (a space-filling Latin hypercube) rather than a
full grid: with four parameters a full grid explodes combinatorially, and a
space-filling design covers the space far more efficiently. Ranges:

- `max_tokens`: 500–2000 terms
- `num_tokens` (n-gram length): 1–2
- `learn_rate`: 0.01–0.1 (log scale)
- `tree_depth`: 4–10

`GRID_SIZE` controls how many configurations are tried. It's kept modest so
the notebook finishes in a few minutes; bump it up (and widen the ranges
above) for a more exhaustive search when you have the compute.

```{r}
#| label: search-space
GRID_SIZE <- 8L

params <- extract_parameter_set_dials(tune_wf) |>
  update(
    max_tokens = max_tokens(range = c(500L, 2000L)),
    num_tokens = num_tokens(range = c(1L, 2L)),
    learn_rate = learn_rate(range = c(-2, -1)),
    tree_depth = tree_depth(range = c(4L, 10L))
  )

set.seed(42)
xgb_grid <- grid_space_filling(params, size = GRID_SIZE, type = "latin_hypercube")
xgb_grid
```

## Cross-validated tuning

Tuning XGBoost + TF-IDF over the full ~40k-row training set is expensive
(`textrecipes` re-tokenizes for every distinct TF-IDF configuration), so we
tune on a stratified subsample and then refit the chosen configuration on the
**full** training set. Increase `TUNE_SAMPLE_N` (up to `nrow(train_df)`) for a
more thorough search when you have the compute/time.

XGBoost itself runs multi-threaded (`nthread` is set on the engine above),
which is where the real speedup comes from; we deliberately don't add an
outer parallel backend on top, since nesting the two tends to oversubscribe
cores without helping.

The folds reuse `cv_folds()` (seed 42, stratified 5-fold) so the evaluation is
consistent with the rest of the project. Results are cached to
`models/tune_tfidf_xgb.rds` so a re-render doesn't re-run the search.

::: {.callout-tip}
## Quick render

The bigram search (`num_tokens = 2`) over the full 5-fold CV is what makes this
chunk slow — building and filtering the bigram vocabulary is recomputed for
every fold/config. On a laptop this can take 10+ minutes. To render quickly,
dial the budget down:

- set `GRID_SIZE <- 5` (in the *Search space* chunk),
- set `TUNE_SAMPLE_N <- 1200` (below), and
- restrict the n-gram range to unigrams by changing `num_tokens` to
  `num_tokens(range = c(1L, 1L))` in the *Search space* chunk.

Unigram features alone reach ~0.99 ROC-AUC on this dataset, so the quick path
costs almost nothing in accuracy. Delete the cached `models/tune_tfidf_xgb.rds`
first if you change these settings, otherwise the old cached run is reused.
:::

```{r}
#| label: tuning
TUNE_SAMPLE_N <- 2500L

tune_train <- if (TUNE_SAMPLE_N < nrow(train_df)) {
  set.seed(42)
  train_df |>
    group_by(generated) |>
    slice_sample(n = ceiling(TUNE_SAMPLE_N / 2)) |>
    ungroup()
} else {
  train_df
}

folds <- cv_folds(tune_train)

tune_path <- here::here("models", "tune_tfidf_xgb.rds")

if (file.exists(tune_path)) {
  tune_res <- readRDS(tune_path)
  message("Loaded cached tuning results from ", tune_path)
} else {
  tune_res <- tune::tune_grid(
    tune_wf,
    resamples = folds,
    grid      = xgb_grid,
    metrics   = cv_metrics(),
    control   = control_grid(event_level = "second", save_pred = FALSE)
  )
  saveRDS(tune_res, tune_path)
  message("Tuning complete; cached to ", tune_path)
}
```

### Tuning results

```{r}
#| label: tuning-results
collect_metrics(tune_res) |>
  filter(.metric == "roc_auc") |>
  arrange(desc(mean)) |>
  select(max_tokens, num_tokens, learn_rate, tree_depth, mean, std_err) |>
  head(10)
```

```{r}
#| label: tuning-plot
#| fig-width: 9
#| fig-height: 6
autoplot(tune_res, metric = "roc_auc") +
  labs(title = "XGBoost + TF-IDF tuning (CV ROC-AUC)")
```

```{r}
#| label: best-params
best_params <- select_best(tune_res, metric = "roc_auc")
best_params
```

## Fit the final model on the full training set

```{r}
#| label: final-fit
final_wf <- finalize_workflow(tune_wf, best_params)

final_fit <- fit(final_wf, data = train_df)
```

## Final evaluation on the held-out test set

This is the **only** time the test set is used.

```{r}
#| label: test-eval
report <- test_set_report(final_fit, test_df)
report$metrics
```

## Confusion matrix analysis

```{r}
#| label: confusion-matrix
fig_dir <- here::here("reports", "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

cm <- report$confusion_matrix
cm

cm_plot <- autoplot(cm, type = "heatmap") +
  labs(title = "Confusion matrix — tuned XGBoost, held-out test set")

ggplot2::ggsave(
  filename = file.path(fig_dir, "confusion_matrix_tuned.png"),
  plot = cm_plot, width = 5, height = 4, dpi = 150
)

cm_plot
```

The table below derives precision, recall, specificity, and the error
breakdown directly from the confusion matrix, treating **AI as the positive
class**.

```{r}
#| label: cm-breakdown
cm_counts <- cm$table
tp <- cm_counts["ai", "ai"]
fp <- cm_counts["ai", "human"]
fn <- cm_counts["human", "ai"]
tn <- cm_counts["human", "human"]

tibble(
  metric = c("True positives (AI called AI)",
             "False positives (human called AI)",
             "False negatives (AI called human)",
             "True negatives (human called human)",
             "Precision (AI)", "Recall / sensitivity (AI)",
             "Specificity (human)", "Accuracy"),
  value = c(tp, fp, fn, tn,
            round(tp / (tp + fp), 4),
            round(tp / (tp + fn), 4),
            round(tn / (tn + fp), 4),
            round((tp + tn) / (tp + fp + fn + tn), 4))
)
```

The paragraph below is generated from the rendered numbers, so it stays in
sync with whatever the model actually produced — copy it straight into the
report.

```{r}
#| label: cm-narrative
#| output: asis
test_acc <- report$metrics |> filter(.metric == "accuracy") |> pull(.estimate)
test_auc <- report$metrics |> filter(.metric == "roc_auc")  |> pull(.estimate)
cv_auc   <- collect_metrics(tune_res) |> filter(.metric == "roc_auc") |>
  slice_max(mean, n = 1) |> pull(mean)
prec <- tp / (tp + fp)
rec  <- tp / (tp + fn)
dom_err <- if (fp >= fn) {
  sprintf("false positives (%d human essays flagged as AI) slightly outnumber the false negatives (%d AI essays missed)", fp, fn)
} else {
  sprintf("false negatives (%d AI essays missed) slightly outnumber the false positives (%d human essays flagged as AI)", fn, fp)
}
gap <- cv_auc - test_auc

cat(sprintf(
"On the held-out test set the final model reaches **%.1f%% accuracy** and a **ROC-AUC of %.3f**, with precision %.3f and recall %.3f for the AI class. Among its mistakes, %s — relevant because for an AI-detection tool a false positive (wrongly accusing a human author) is usually the costlier error. The cross-validation ROC-AUC during tuning was %.3f, so the test-set score is %s the CV estimate (a gap of %+.3f), which gives %s that the model is not badly overfit to the tuning folds.",
100 * test_acc, test_auc, prec, rec, dom_err, cv_auc,
ifelse(gap > 0, "below", "above"), -gap,
ifelse(abs(gap) < 0.02, "reassurance", "some cause for caution")
))
```

## Did tuning help? (vs. the untuned comparison model)

```{r}
#| label: compare-untuned
# Baseline: the recipe/model settings used in 04-model-comparison.qmd
baseline_rec <- tfidf_recipe(generated ~ text, data = train_df, max_tokens = 1000)
baseline_spec <- boost_tree(trees = 300, learn_rate = 0.05) |>
  set_engine("xgboost", nthread = parallel::detectCores()) |>
  set_mode("classification")

baseline_fit <- workflow() |>
  add_recipe(baseline_rec) |>
  add_model(baseline_spec) |>
  fit(data = train_df)

baseline_report <- test_set_report(baseline_fit, test_df)

bind_rows(
  baseline_report$metrics |> mutate(model = "untuned (nb 04 settings)"),
  report$metrics          |> mutate(model = "tuned (this notebook)")
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = model, values_from = .estimate)
```

```{r}
#| label: tuning-gain
#| output: asis
base_auc  <- baseline_report$metrics |> filter(.metric == "roc_auc") |> pull(.estimate)
tuned_auc <- report$metrics          |> filter(.metric == "roc_auc") |> pull(.estimate)
delta <- tuned_auc - base_auc
cat(sprintf(
"Tuning the TF-IDF representation and XGBoost jointly changed test-set ROC-AUC by %+.3f (from %.3f untuned to %.3f tuned). %s",
delta, base_auc, tuned_auc,
if (delta > 0.002) "The joint search delivered a measurable improvement." else
if (delta < -0.002) "The untuned defaults were already competitive here; the gain from tuning is negligible, which itself is worth reporting." else
"Both are essentially tied, suggesting the problem is easy enough that the default settings already capture most of the signal."
))
```

## Persist the final tuned model

```{r}
#| label: save-model
models_dir <- here::here("models")
dir.create(models_dir, showWarnings = FALSE, recursive = TRUE)
saveRDS(final_fit, file = file.path(models_dir, "xgboost_tuned_final.rds"))
```

## Notes for the report

- **Headline result:** state the final held-out test metrics (accuracy and
  ROC-AUC) as the top-line finding. As a reference point, a single untuned
  unigram XGBoost already reaches ~0.94 accuracy / ~0.99 ROC-AUC on this test
  set, so the task is highly separable — frame the tuned model as confirming
  and refining that rather than as a dramatic jump.
- **Winning hyperparameters:** record `best_params` and explain the trade-offs
  — did a larger vocabulary or adding bigrams (`num_tokens = 2`) actually help,
  or did the simpler representation win? If unigrams win, that's a clean,
  defensible result.
- **Which mistakes:** use the confusion-matrix breakdown to discuss the error
  profile (false positives vs. false negatives) and connect it to the
  fair-evaluation / overfitting concerns from the project memo.
- **Tie back to EDA:** the most informative TF-IDF features (`vip::vi()` above)
  should echo the EDA in `01-eda.qmd` — e.g., AI text leaning on more abstract,
  technical vocabulary ("model", "data", "system") and humans on more everyday
  words. If the top features line up with the descriptive findings, that's
  evidence the model is learning genuine linguistic differences rather than
  dataset artifacts.
```
